Behind Bars: Insights into the Wellbeing of U.S. Prisoners

### 2024-02-16 paper/paper.qmd
included the introduction with updated references

>>>>

The rising death rates among prisoners in the challenging environment of American jails show discrepancies, difficulties,
and institutional failings. Issues including institutional prejudice and poor healthcare have been brought to light by
a recent news report. For instance, the article talks about a study which discovered that although 0.85% of Americans
with asthma were incarcerated, just 0.15% of asthma patients received treatment, an inequality that was more than five
times higher (Murez_2023). Our goals are to identify the causes of the growing death rate, recognize the influence of
demographics and highlight the significant negative effects of prison circumstances on the mental health of prisoners.

According to the US Justice Department's Office of the Inspector General (OIG), 187 individuals died by suicide inside
Federal Bureau of Prisons(BoP) institutions between 2014 and 2021. The staff providing psychiatric services in the prisons
stated that these deaths could have been avoided if the facilities followed protocol and provided inmates with the necessary
resources and treatment. In addition, the study listed 56 "accidental" fatalities and 89 homicides that occurred during that
time period. It also claimed that the BoP often neglected to appropriately punish employees for actions that led to the deaths.(Levin 2024)
News reports highlight the critical need for a full understanding of prisoners' problems, going beyond simple statistics.
This has sparked calls for reform and raised concerns about the efficacy of healthcare in prisons.

Homicides accounted for the second-highest number of fatalities recorded in the survey, with one prisoner, Bulger, being
beaten to death by other inmates in 2018 (Associated Press_2024). Editorials raise ethical concerns regarding the prison
system by arguing that inequality in healthcare and the high rate of avoidable diseases among prisoners are reflections
of larger social issues. Investigative articles highlight patterns of prejudice and neglect while exposing the realities
of minorities within jails.

As we dig deeper, our focus goes beyond statistics, aiming to amplify the voices and experiences of those directly
affected. This exploration goes beyond the numbers, delving into the personal dimensions of the prison experience and
laying the groundwork for a comprehensive analysis that seeks to drive change in a system facing significant challenges.
